---
title: "Homework 3: The effect of the study variables"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(metafor)
knitr::opts_chunk$set(echo = FALSE, fig.width = 5, fig.height = 5, fig.align = 'center')

dat.bcg.transformed <- escalc(
  measure = 'RR',  ## the outcome measure to use, RR -> (log) risk ratio
  ai = tpos,       ## upper left, of the 2 x 2 frequency table
  bi = tneg,       ## upper right, ...
  ci = cpos,       ## lower left, ...
  di = cneg,       ## lower right, ...
  data = dat.bcg
)

res <- rma(yi, vi, data = dat.bcg.transformed)
res2 <- rma(yi, vi, mods = ~ ablat, data = dat.bcg.transformed)
```

In the practice problem, we used the random-effects model to get an estimate of
the overall efficacy of the BCG vaccine across different studies.

The following code re-applies the mode and shows the results.

```{r recape}
# load library
library(metafor)

# calculate outcome measure
dat.bcg.transformed <- escalc(
  measure = 'RR',  ## the outcome measure to use, RR -> (log) risk ratio
  ai = tpos,       ## upper left, of the 2 x 2 frequency table
  bi = tneg,       ## upper right, ...
  ci = cpos,       ## lower left, ...
  di = cneg,       ## lower right, ...
  data = dat.bcg
)

# apply the random-effects model
res <- rma(yi, vi, data = dat.bcg.transformed)
res
```

Based on the model above we rejected the null hypothesis stating that the 
studies are not heterogeneous. We also got an estimate of the contribution of
the variability between the studies to the outcome measure of around 92%. 

The original dataset contains two variables that might contribute to the 
different outcomes of the studies. These are

- `ablat`: absolute latitude of the study location (in degrees)
- `alloc`: method of treatment allocation (random, alternate, or systematic)

We want to include these variables in a model similar to the above and estimate
their contributions to the outcome and test whether they are significant or not.

To do that, you will use in addition to the code above the argument `mod`. This
arguments takes a formula of the additional variables to include in the model.
The resulting output will estimate the contribution of each variable and tests
for its significance.

```{r mixed, exercise=TRUE}
# apply the mixed effects model
```

```{r mixed-solution}
# apply the mixed effects model
res2 <- rma(yi, vi, mods = ~ ablat + alloc, data = dat.bcg.transformed)
res2
```

Your code should return the results of the mixed-effects model that include
which will help you answer the following questions.

```{r quiz}
quiz(
  caption = 'Quiz',
  question(
    text = 'Which of the two variables have a significant effect on the vaccine efficacy?',
    answer('Tha absolute lattiude of the location', correct = TRUE),
    answer('The allocation method of the treatment'),
    random_answer_order = TRUE
  ),
  question(
    text = 'How much change in the log relative risk is precipitated by one degree increase in the absoloute lattitude of the study location?',
    answer('0.3'),
    answer('0.03'),
    answer('-0.03', correct = TRUE),
    random_answer_order = TRUE
  )
)
```

The following graph should confirm the you results obtained earlier. It shows 
in visual terms the correlation between the outcome measure and one of the two
study variables.

```{r bubbles}
# visualize the correlation between ablat and risk ratio
regplot(res2, mod = 'ablat')
```
