---
title: "Evaluating the effects of the BCG vaccine"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(metafor)
knitr::opts_chunk$set(echo = FALSE, fig.width = 5, fig.height = 5, fig.align = 'center')

dat.bcg.transformed <- escalc(
  measure = 'RR',  ## the outcome measure to use, RR -> (log) risk ratio
  ai = tpos,       ## upper left, of the 2 x 2 frequency table
  bi = tneg,       ## upper right, ...
  ci = cpos,       ## lower left, ...
  di = cneg,       ## lower right, ...
  data = dat.bcg
)
res <- rma(yi, vi, data = dat.bcg.transformed)

```

## What is a meta-analysis? 

A meta-analysis is a statistical analysis that combines the results of multiple
scientific studies. The goal of this document is to present an example of such
analysis. 

We will achieve this in a few steps

- Loading and exploring a dataset of published studies
- Transforming the main effects and variance to an appropriate format
- Applying a random-effects model to the dataset
- Exploring and visualizing the output of the model

These materials are based on the 

- The [vignette](https://www.metafor-project.org/) of the `metafore` package
- The article by [Viechtbauer, 2010](https://www.jstatsoft.org/article/view/v036i03) 

---

## Loading the BCG dataset

The following analysis uses the `dat.bcg` dataset and the `metafor` R package.
The goal is to estimate the efficacy of the BCG vaccine and to relate its effect
to study variables such as the latitude of the study location and the method
of treatment allocation.

The dataset `dat.bcg` is shipped with the `metafor` package. So we first need to
load the package and the dataset would be available to us.

```{r load, echo=TRUE}
# load the package
library(metafor)
```

Once you the package has been loaded, the dataset will be available to you.
Go on and excute the following code to examine the first three rows.

```{r head, exercise=TRUE}
# examine the dataset
head(dat.bcg, 3)
```

This dataset consists of `r nrow(dat.bcg)` clinical trials. Each trial had a few
variables recorded in addition to `author` and `year`. These are

- `tpos`: number of TB positive cases in the treated (vaccinated) group
- `tneg`: number of TB negative cases in the treated (vaccinated) group
- `cpos`: number of TB positive cases in the control (non-vaccinated) group
- `cneg`: number of TB negative cases in the control (non-vaccinated) group
- `ablat`: absolute latitude of the study location (in degrees)
- `alloc`: method of treatment allocation (random, alternate, or systematic)

To make sure you understand how the data from these studies in encoded, have a 
look at the table below, and answer the questions

|                | positive | negative |
|----------------|----------|----------|
| vaccinated     |     a    |     b    |
| not-vaccinated |     c    |     d    |

```{r quiz-1}
quiz(
  caption = 'Quiz 1',
  question(
    text = 'In the table above, a indicates?',
    answer('The number of people who recieved the vaccine and tested positive', correct = TRUE),
    answer('The number of people who recieved the vaccine'),
    answer('The number of people who tested positive'),
    random_answer_order = TRUE
  ),
  question(
    text = 'What does the data in the table above represent?',
    answer('The observations in the individual studies', correct = TRUE),
    answer('The total number of observations across the studies'),
    answer('The averaged observations accross the studies'),
    random_answer_order = TRUE
  ),
  question(
    text = 'What is the number of participants in the first three studies?',
    answer('262, 609, and 451', correct = TRUE),
    answer('123, 306, and 231'),
    answer('139, 303, and 220'),
    random_answer_order = TRUE
  )
)
```

The data presented in this way may be easy to read for the individual studies.
However, to be able to aggregate and compare the results across the studies,
further transformation of the data will be needed. We turn to this next.

----

## Calculating the outcome measure

To be able to conduct a meta-analysis of the study, we first need to do some
transformations on the dataset. These are

1. Calculate the log risk ratio for each study
2. Calculate the sampling variance for each study

The choice of which outcome measure (effect size) to use depends on the goal of
the analysis, the design of the studies and the available information.

Fortunately, the `metafor` package provides a convenient function, `escalc` to
calculate the common outcome measures.

The following code takes the content of the individual 2 by 2 tables and 
calculates the risk ratio (RR) on the log scale and the the sampling variance.

```{r transform, echo=TRUE}
# transform
dat.bcg.transformed <- escalc(
  measure = 'RR',  ## the outcome measure to use, RR -> (log) risk ratio
  ai = tpos,       ## upper left, of the 2 x 2 frequency table
  bi = tneg,       ## upper right, ...
  ci = cpos,       ## lower left, ...
  di = cneg,       ## lower right, ...
  data = dat.bcg
)
```

The resulting object (`dat.bcg.transformed`) is the same as before, with added 
two columns. Examine the object below and figure out the names and meanings of 
the new columns. 

```{r transformed,exercise=TRUE}
# examine the returned object 

# the added new columns
```

```{r transformed-solution}
# examine the returned object 
dat.bcg.transformed

# the added new columns
setdiff(names(dat.bcg.transformed), names(dat.bcg))
```

From the information and the code above, you should be able to address the 
following questions.

```{r quiz-2}
quiz(
  caption = 'Quiz 2',
  question(
    text = 'What are the names of the new columns in "dat.bcg.transformed"?',
    answer('yi and vi', correct = TRUE),
    answer('ablat and alloc'),
    answer('m1 and m2'),
    random_answer_order = TRUE
  ),
  question(
    text = 'The new columns in "dat.bcg.transformed" contains, in left to right order?',
    answer('Effect size of the outcome and sampleing variance of the study', correct = TRUE),
    answer('Sampleing variance of the study and effect size of the outcome'),
    random_answer_order = TRUE
  ),
  question(
    text = 'What is the relative risk of vaccinated to non-vaccinated people in Aronson, 1948 to test positive for TB?',
    answer('0.89 less likely', correct = TRUE),
    answer('0.89 more likely'),
    answer('0.33 less likely'),
    random_answer_order = TRUE
  )
)
```

Now that we got the dataset in shape, we will turn to modeling the aggregate
effect of the vaccine using a random-effects model in the next section.

----

## Applying a random-effects model

The `metafor` package provides a unified interface for the meta-analytic linear
regression framework. The function we will be using is called `rma`. You can 
read more about this function by calling `?rma` to get the help page.

Because the included studies are not identical, we need a way to deal with the 
variability (heterogeneity) introduced by the difference between the studies.
The particular model we will use here is the "Random-effects model." This model
is used when the heterogeneity between the studies is unobserved but is 
independent from the study variables. This is in contrast to the the "fixed"
effects where the heterogeneity is assumed to be correlated with the study 
variables.

Briefly, we assume for $i = 1, ..., k$ studies

$$
y_i=\theta_i + e_i\\
\theta_i=\mu + u_i\\
e_i \sim \text{Normal}(0, v_i)\\
u_i \sim \text{Normal}(0, \tau^2)
$$

where in each study $i$,

- $y_i$ is the observed effect size
- $\theta_i$ the true (unknown) effect size
- $e_i$ is the sampling error

If we treat heterogeneity as random, then

- $e_i$ will be randomly distributed and depends on the study variance $v_i$ 
- the true effects $u_i$ will also be randomly distributed and depend on $\tau^2$.

The goal would be to estimate

- $\mu$, the average true effect
- $\tau^2$, the total amount of heterogeneity among the true effects

If $\tau^2=0$ the the true effects are homogeneous and $\mu = \theta$.


Applying the model is way easier than reading it. We pass the effect size `yi`,
the variance `vi` column names to and the data `dat.bcg.transformed` to `rma`
function.

```{r model, echo=TRUE}
# apply random-effects model
res <- rma(yi, vi, data = dat.bcg.transformed)
res
```

We performed these calculations at the log scale, so we need to transform the
output to the natural scale which easier to comprehend. Go ahead and excute the 
code and take note of the output.

```{r predict, exercise=TRUE}
# return to natural scale from log
predict(res, transf=exp, digits=2)
```

From the results above try answering the following questions.

```{r quiz-3}
quiz(
  caption = 'Quiz 3',
  question(
    text = "According to the model, which statements are true?",
    answer('a tuberculosis infection in vaccinatedindividuals is on average half as large as the infection risk without the vaccination', correct = TRUE),
    answer('92% of the total variablility in the estimate can be attributed to heterogeneity between the studies', correct = TRUE),
    answer('The ratio between the variablility between studies to sampling variability is 13:1', correct = TRUE),
    answer('Saying that there is no heterogeneity between studies should be rejected with p-value < .001.', correct = TRUE),
    random_answer_order = TRUE
  )
)
```

## Exploring and visualizing the output

When the number of studies is small, and the effects are mostly in one direction
, it would be easy to read and interpret the results from the text output.
However, visualizing the output would become handy with larger numbers of 
studies and when the effects are mixed. Visualization also help explore 
different aspects of the data that are not immediatly obvious in text.

We will try out two types of plots that are standard in meta-analysis reports.

### The forest plot

Forest plots show the estimated effect size of the individual studies along with
the overall estimate from the linear model.

The `metafor` package provides a convenient function to produce this type of 
graph called `forest`. There are many options you could use to customize the
output, but we will try it out using the default options.

```{r forest}
# make a forest plot
forest(res)
```

The figure shows the log risk ratio of the individual studies along with the 
confidence intervals. At the bottom is the estimated overall effect from the
random-effects model.

Interpreting this plot is fairly straight forward. Have a look and answer the
following questions.

```{r quiz-4}
quiz(
  caption = 'Quiz 4',
  question(
    text = 'Which study has an outcome in the opposing direction to all others?',
    answer('Study 11'),
    answer('Study 12', correct = TRUE),
    answer('Study 13'),
    random_answer_order = TRUE
  ),
  question(
    text = 'What is the lower confidence interval of the estimated overall effect from the random effects model?',
    answer('-0.36'),
    answer('-0.71'),
    answer('-1.07', correct = TRUE),
    random_answer_order = TRUE
  )
)
```

### The funnel plot

Funnel plots are another handy way to visualize the effect sizes and their 
sampling errors. The first is displayed on the horizontal axis and the second on 
the vertical axis. Other lines are draws to indicate the estimate based on the 
model and the confidence interval around it.

Here is the code to produce the graph using the `funnel` function from `metafore`

```{r funnel}
# make a funnel plot
funnel(res)
```

Notice that we only used the default options, there are several ways to 
customize the plot and overlay more information. 

```{r quiz-5}
question(
    text = 'How many studies fall outside the estimated (pseudo) confidence interval?',
    answer('5'),
    answer('6', correct = TRUE),
    answer('7'),
    random_answer_order = TRUE
  )
```

These and other visualization can be useful tools to diagnose problems with the
data or the analysis. For example, it is clear from the previous plots that 
there is large variability among the studies, some of which might be accounted
for using the other study variables.

In the homework problem, you will be asked to attempt accounting for these 
variables to improve the overall estimate of the outcomes.
